### 2022-07-19 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@587d1042f4ada9e8e096c69a8f3032b236224fdd 🚀

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\thispagestyle{empty}
{\Large\bfseries\sffamily Maximum entropy modelling of food-web
structure}
\vskip 5em

@@

\vfill
\textbf{\sffamily Abstract: }The principle of maximum entropy is a
rigorous mathematical method of finding constrained probability
distributions that has been proven useful in many ecological modelling
problems. However, despite its broad application in graph and network
theory, it has seldom been used to model ecological networks. Here we
show how the (joint) degree distribution of maximum entropy can be
directly derived using the number of species and the number of links in
terrestrial and aquatic food webs. We also present a heuristic and
flexible approach of finding the network of maximum entropy based on
simulating annealing and SVD entropy. We built two of these
network-level models using constraints given by the connectance and the
joint degree sequence, respectively. All maximum entropy models were
compared against open access food-web data and null and neutral models
commonly used in network ecology. We found that the maximum entropy
network model constrained by the joint degree sequence was a good
predictor of food-web structure, including nestedness and motifs
distribution. Overall, our results suggest that many properties of
ecological networks are mainly driven by the joint degree distribution
and statistical phenomena.
\vfill

@@

We acknowledge that this study was conducted on land within the
traditional unceded territory of the Saint Lawrence Iroquoian,
Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak nations. This
work was supported by the Institute for Data Valorisation (IVADO) and
the NSERC BIOS\(^2\) CREATE program.

>>>>

\fancypagestyle{normal}
{
  \fancyhf{}
  \fancyfoot[R]{\footnotesize\sffamily\thepage\ of \pageref*{LastPage}}
}
\begin{document}
\thispagestyle{empty}
{\Large\bfseries\sffamily What constrains food webs? A maximum entropy
framework for predicting their structure with minimal biases}
\vskip 5em

@@

\vfill
\textbf{\sffamily Abstract: }Food webs are complex ecological networks
whose structure is both ecologically and statistically constrained, with
many network properties being correlated with each other. Despite the
recognition of these invariable relationships in food webs, the use of
the principle of maximum entropy (MaxEnt) in network ecology is still
rare. This is surprising considering that MaxEnt is a renowned and
rigorous statistical tool precisely designed for understanding and
predicting many different types of constrained systems. Precisely, this
principle asserts that the least-biased probability distribution of a
system's property, constrained by prior knowledge about that system, is
the one with maximum information entropy. MaxEnt has been proven useful
in many ecological modelling problems, but its application in food webs
and other ecological networks is limited. Here we show how MaxEnt can be
used to derive many food-web properties both analytically and
heuristically. First, we show how the joint degree distribution (the
joint probability distribution of the numbers of prey and predators for
each species in the network) can be derived analytically using the
number of species and the number of interactions in food webs. Second,
we present a heuristic and flexible approach of finding a network's
adjacency matrix (the network's representation in matrix format) based
on simulated annealing and SVD entropy. We built two heuristic models
using the connectance and the joint degree sequence as statistical
constraints, respectively. We compared both models' predictions against
corresponding null and neutral models commonly used in network ecology
using open access data of terrestrial and aquatic food webs sampled
globally (N = 257). We found that the heuristic model constrained by the
joint degree sequence was a good predictor of many measures of food-web
structure, especially the nestedness and motifs distribution.
Specifically, our results suggest that the structure of terrestrial and
aquatic food webs is mainly driven by their joint degree distribution.
\vfill

@@

We acknowledge that this study was conducted on land within the
traditional unceded territory of the Saint Lawrence Iroquoian,
Anishinabewaki, Mohawk, Huron-Wendat, and Omàmiwininiwak nations. This
work was supported by the Institute for Data Valorisation (IVADO) and
the Natural Sciences and Engineering Research Council of Canada (NSERC)
Collaborative Research and Training Experience (CREATE) program, through
the Computational Biodiversity Science and Services (BIOS\(^2\))
program.

### 2022-07-19 preprint.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@587d1042f4ada9e8e096c69a8f3032b236224fdd 🚀

\noindent {\LARGE \textbf{\textsf{Maximum entropy modelling of food-web
structure}}}

@@

\textbf{Abstract}:\,The principle of maximum entropy is a rigorous
mathematical method of finding constrained probability distributions
that has been proven useful in many ecological modelling problems.
However, despite its broad application in graph and network theory, it
has seldom been used to model ecological networks. Here we show how the
(joint) degree distribution of maximum entropy can be directly derived
using the number of species and the number of links in terrestrial and
aquatic food webs. We also present a heuristic and flexible approach of
finding the network of maximum entropy based on simulating annealing and
SVD entropy. We built two of these network-level models using
constraints given by the connectance and the joint degree sequence,
respectively. All maximum entropy models were compared against open
access food-web data and null and neutral models commonly used in
network ecology. We found that the maximum entropy network model
constrained by the joint degree sequence was a good predictor of
food-web structure, including nestedness and motifs distribution.
Overall, our results suggest that many properties of ecological networks
are mainly driven by the joint degree distribution and statistical
phenomena.

>>>>

\noindent {\LARGE \textbf{\textsf{What constrains food webs? A maximum
entropy framework for predicting their structure with minimal biases}}}

@@

\textbf{Abstract}:\,Food webs are complex ecological networks whose
structure is both ecologically and statistically constrained, with many
network properties being correlated with each other. Despite the
recognition of these invariable relationships in food webs, the use of
the principle of maximum entropy (MaxEnt) in network ecology is still
rare. This is surprising considering that MaxEnt is a renowned and
rigorous statistical tool precisely designed for understanding and
predicting many different types of constrained systems. Precisely, this
principle asserts that the least-biased probability distribution of a
system's property, constrained by prior knowledge about that system, is
the one with maximum information entropy. MaxEnt has been proven useful
in many ecological modelling problems, but its application in food webs
and other ecological networks is limited. Here we show how MaxEnt can be
used to derive many food-web properties both analytically and
heuristically. First, we show how the joint degree distribution (the
joint probability distribution of the numbers of prey and predators for
each species in the network) can be derived analytically using the
number of species and the number of interactions in food webs. Second,
we present a heuristic and flexible approach of finding a network's
adjacency matrix (the network's representation in matrix format) based
on simulated annealing and SVD entropy. We built two heuristic models
using the connectance and the joint degree sequence as statistical
constraints, respectively. We compared both models' predictions against
corresponding null and neutral models commonly used in network ecology
using open access data of terrestrial and aquatic food webs sampled
globally (N = 257). We found that the heuristic model constrained by the
joint degree sequence was a good predictor of many measures of food-web
structure, especially the nestedness and motifs distribution.
Specifically, our results suggest that the structure of terrestrial and
aquatic food webs is mainly driven by their joint degree distribution.
